\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{hyperref}
\usepackage{enumitem}

\hypersetup{
  colorlinks=true,
  linkcolor=black,
  urlcolor=blue
}

\setlist[itemize]{leftmargin=*}

\title{IFT GPU Workstation Memo}
\author{CQMP}
\date{\today}

\begin{document}
\maketitle

\section{Hardware Configuration}

The workstations are Dell Pro Max Tower T2 systems installed from the shared
workstation image.  The currently installed reference machine, Matsubara, has
the following configuration:

\begin{itemize}
  \item CPU: Intel Core Ultra 7 265, 20 cores, one socket.
  \item Memory: 30 GiB RAM plus 8 GiB compressed swap.
  \item Local storage: one Kioxia XG10d 1 TB NVMe device.  The standard
        installation uses a 512 MB EFI partition, a 1 GB \texttt{/boot}
        partition, a 50 GB operating-system partition, and the remainder as
        \texttt{/data}.
  \item Graphics: integrated Intel Arrow Lake-S graphics plus an NVIDIA
        RTX 2000 Ada Generation GPU.
  \item Network: Intel I219-LM Ethernet.
\end{itemize}

The workstation inventory and assigned addresses are tracked in
\texttt{machines.md} in the image repository.

\section{System Configuration}

The machines run CentOS Stream 9 as a bootc image.  In this model, the
operating system is built as a container image and deployed atomically to the
host.  The machine normally boots the current deployment and keeps a rollback
deployment available.  Updates are delivered by publishing a new image and then
running a bootc upgrade on the workstation.

The image repository is:

\begin{center}
\url{https://github.com/CQMP/workstation-image}
\end{center}

The published image used by the installer and by bootc is:

\begin{center}
\texttt{ghcr.io/cqmp/centos9-workstation:latest}
\end{center}

The main image definition is the \texttt{Containerfile}.  The unattended
installer is described by \texttt{ks.cfg}; it partitions the local NVMe disk,
pulls the image above, and installs the bootc system.  The image includes GNOME
and GDM, NVIDIA drivers and CUDA tooling, SSH, LDAP/SSSD authentication,
autofs/NFS integration, HTCondor, compilers, MPI stacks, Python scientific
packages, browsers, editors, and common desktop tools.

The image also carries kernel arguments and initramfs configuration needed for
these systems.  In particular, nouveau is blacklisted, NVIDIA DRM mode setting
is enabled, and the initramfs is kept small enough for reliable EFI boot.

\section{Package Requests and Image Changes}

Users should not install long-lived system packages manually on individual
workstations.  Manual changes are local to one machine and may disappear after
reinstallation or diverge from the rest of the pool.  Instead, package requests
and system configuration changes should be made through the image repository.

The preferred workflow is:

\begin{enumerate}
  \item Fork \url{https://github.com/CQMP/workstation-image}.
  \item Modify the \texttt{Containerfile} or the relevant file under
        \texttt{etc/}, \texttt{usr/}, or another tracked directory.
  \item Build or inspect the change locally when practical.
  \item Submit a pull request against the main repository.
\end{enumerate}

After a pull request is merged, GitHub Actions rebuilds and publishes the image.
Machines can then receive the update with:

\begin{verbatim}
sudo bootc upgrade
sudo reboot
\end{verbatim}

\section{Logging In}

Authentication is handled through the FUW/ITP LDAP system via SSSD.  Users log
in with their Unix password.  Being present in LDAP is necessary but not
sufficient: the image also contains an explicit allow list of users who may log
in to these workstations.

The allow list is configured in \texttt{etc/sssd/sssd.conf} as
\texttt{simple\_allow\_users}.  Users who need access should provide their Unix
user name so it can be added to the image.  The current image allow list is:

\begin{center}
\texttt{egull, rfarid, ajazdzewska, amarie, abalbi}
\end{center}

Home directories and shared file systems are resolved through LDAP/autofs.  If
login fails, the first things to check are whether the user exists in the LDAP
Unix domain, whether the user is on the image allow list, and whether the user
is using the Unix password rather than another institutional password.

\section{Condor}

Each workstation is configured as an HTCondor execute node.  The local Condor
configuration is in:

\begin{center}
\texttt{etc/condor/config.d/00-ift-execute.conf}
\end{center}

The execute node contacts \texttt{condor.gull-group.org}.  It runs only the
Condor master and startd daemons, so these machines provide execution slots but
are not the central manager.

The configuration is designed for interactive workstations.  Jobs start only
when the keyboard has been idle for at least 15 minutes and the machine load is
low:

\begin{center}
\texttt{START = (KeyboardIdle > 900) \&\& (LoadAvg < 0.5)}
\end{center}

If a local user returns briefly, Condor suspends the job.  If the machine stays
active, the job is preempted and evicted.  The image enables CRIU-based
checkpointing and installs NVIDIA's \texttt{cuda-checkpoint} helper so GPU jobs
can be checkpointed during eviction when supported by the workload.  The
configured vacate time is one hour, giving checkpointing time to complete before
the job is killed.

The NVIDIA GPU is intended to be available to CUDA and Condor jobs.  Users
submitting GPU jobs should request the appropriate GPU resources through the
central Condor configuration and should test their jobs with \texttt{nvidia-smi}
or CUDA device discovery before launching large runs.

\section{File System}

Authentication and network file-system access are handled through ITP/FUW Unix
credentials.  The image configures SSSD for LDAP and autofs for the shared
mount points:

\begin{center}
\texttt{/dmj}, \texttt{/expo}, and \texttt{/repo}
\end{center}

These network file systems are appropriate for home directories, shared code,
small inputs, and files that need central management.

Each workstation also has a fast local NVMe scratch disk mounted at
\texttt{/data}.  The standard installation creates per-user directories there
for allowed users.  This disk is not backed up.  It should be used for
simulation scratch data, compilation trees, temporary files, local virtual
environments, and other high-I/O work that does not need backup.  Important
results should be copied back to backed-up network storage.

\section{Remote Desktop Access (VNC)}

The workstations run TigerVNC server (\texttt{tigervnc-server}, EL9). Each user
runs on a chosen display number (e.g., \texttt{:1} on port 5901), managed as a
\emph{system-level} systemd service via a display-to-user mapping file --- there
is no per-user \texttt{systemctl --user} template on this package version.
Sessions persist across disconnects: GUI applications keep running after the
viewer is closed.

If two users share a machine they must use different display numbers
(\texttt{:1}, \texttt{:2}, \ldots); the corresponding TCP port is
$5900 + \text{display number}$.

\subsection*{First-time setup (on the workstation, via SSH)}

Set a VNC password (as yourself, not root):
\begin{verbatim}
vncpasswd
\end{verbatim}

As root, map your display number to your username:
\begin{verbatim}
echo ":1=<username>" >> /etc/tigervnc/vncserver.users
\end{verbatim}

As root, enable and start the service (no \texttt{--user} flag --- this is a
system unit):
\begin{verbatim}
systemctl enable --now vncserver@:1.service
\end{verbatim}

The session will now start automatically after every reboot. The default
session type is GNOME, set system-wide in
\texttt{/etc/tigervnc/vncserver-config-defaults}.

\subsection*{Connecting from macOS}

VNC traffic is unencrypted; always route it through an SSH tunnel.  In a
terminal, forward the VNC port to your local machine:

\begin{verbatim}
ssh -N -L 5901:Matsubara:5901 <username>@<machine-ip>
\end{verbatim}

Keep that terminal open (or background it with \texttt{\&}).  Then open the
built-in macOS Screen Sharing app---press \textbf{Cmd+K} in Finder, or
search for ``Screen Sharing'' in Spotlight---and connect to:

\begin{verbatim}
vnc://localhost:5901
\end{verbatim}

\href{https://www.realvnc.com/en/connect/download/viewer/}{RealVNC Viewer}
(free download) is a more comfortable alternative: it offers better compression,
a saved-connections list, and clipboard integration.  After the SSH tunnel is
running, create a new connection to \texttt{localhost:5901}.

\subsection*{Session management}

As root:
\begin{verbatim}
systemctl stop    vncserver@:1.service   # stop the session
systemctl disable vncserver@:1.service   # prevent autostart on reboot
\end{verbatim}

\section{Matrix/Element Messaging Server}

\subsection*{Architecture}

The group runs a self-hosted Matrix homeserver to replace the Slack workspace
(which has a 90-day message history limit on the free plan).  Matrix user IDs
take the form \texttt{@user:matrix.gull-group.org}.  The client is Element
Desktop, installed on all workstations.

The deployment spans two machines:

\begin{itemize}
  \item \textbf{Oracle Cloud VM} (\texttt{gull-matrix-vnic}, VM.Standard.E2.1.Micro,
        1 OCPU / 1 GB RAM / 2 GB swap, Ubuntu 24.04 LTS, Frankfurt region).
        Public IP \texttt{79.76.114.255}; DNS A record
        \texttt{matrix.gull-group.org} points here.  Runs Tuwunel (Matrix
        homeserver) and Nginx (TLS reverse proxy).  This VM is on the Oracle
        Cloud Always-Free tier and costs nothing.

  \item \textbf{Synology NAS} (\texttt{fermi.physics.lsa.umich.edu}, DS2419+,
        DSM 7.3.2).  Runs MinIO (S3-compatible object store) in Docker,
        providing durable media storage on local NAS disks.  The Synology is
        behind the University of Michigan campus firewall; it cannot accept
        inbound connections but can call out.
\end{itemize}

Because the Synology is behind a firewall, a persistent \emph{reverse} SSH
tunnel --- maintained by an \texttt{autossh} Docker container on the Synology
--- forwards port 9000 on the Oracle VM's loopback interface through to MinIO
on the Synology.  Tuwunel stores and retrieves all media (files, images,
avatars) over this tunnel.  MinIO is never reachable directly from the
internet.  The full request path is:

\begin{center}
Element \;$\to$\; HTTPS :443 \;$\to$\; Nginx \;$\to$\; Tuwunel :8008 \;$\to$\; tunnel :9000 \;$\to$\; MinIO on Synology
\end{center}

\subsection*{Services and their locations}

\noindent\textbf{On the Oracle VM} (\texttt{ssh ubuntu@79.76.114.255}):

\begin{center}
\begin{tabular}{lll}
\hline
Service & systemd unit & Notes \\
\hline
Tuwunel & \texttt{tuwunel.service} & Listens on \texttt{127.0.0.1:8008} \\
Nginx   & \texttt{nginx.service}   & Listens on :80 (redirect) and :443 (TLS) \\
Certbot & \texttt{certbot.timer}   & Auto-renews TLS cert; next expiry 2026-09-18 \\
\hline
\end{tabular}
\end{center}

\noindent\textbf{On the Synology} (\texttt{fermi.physics.lsa.umich.edu}):

\begin{center}
\begin{tabular}{ll}
\hline
Container & Purpose \\
\hline
\texttt{matrix-minio}  & MinIO object store, \texttt{127.0.0.1:9000} \\
\texttt{matrix-tunnel} & autossh reverse tunnel to Oracle VM \\
\hline
\end{tabular}
\end{center}

Both containers are defined in
\texttt{/volume1/docker/matrix/docker-compose.yml} and restart automatically.

\subsection*{Key file locations}

\begin{center}
\begin{tabular}{ll}
\hline
\multicolumn{2}{l}{\textit{Oracle VM}} \\
\hline
Tuwunel config        & \texttt{/etc/tuwunel/tuwunel.toml} \\
Tuwunel database      & \texttt{/var/lib/tuwunel} (RocksDB) \\
Nginx site config     & \texttt{/etc/nginx/sites-available/matrix} \\
TLS certificate       & \texttt{/etc/letsencrypt/live/matrix.gull-group.org/} \\
SSH tunnel user       & \texttt{/home/tunnel/.ssh/authorized\_keys} \\
SSH tunnel restriction& \texttt{/etc/ssh/sshd\_config} (Match User tunnel block) \\
\hline
\multicolumn{2}{l}{\textit{Synology}} \\
\hline
Docker compose file   & \texttt{/volume1/docker/matrix/docker-compose.yml} \\
MinIO data            & \texttt{/volume1/minio/data} \\
Tunnel SSH key        & \texttt{/volume1/docker/matrix-tunnel/tunnel\_key} \\
\hline
\end{tabular}
\end{center}

\subsection*{Key configuration notes}

\textbf{Tuwunel S3 block} --- MinIO runs over plain HTTP (no TLS) because the
tunnel is already on loopback.  Tuwunel's S3 client defaults to
\texttt{use\_https = true} and will fail with a cryptic ``builder error'' if
this is not overridden.  The working config block in
\texttt{/etc/tuwunel/tuwunel.toml} is:

\begin{verbatim}
media_storage_providers = ["media_on_s3"]
store_media_on_providers = ["media_on_s3"]

[global.storage_provider.media_on_s3.s3]
bucket = "matrix-media"
region = "us-east-1"
key = "matrix"
secret = "<see docker-compose.yml on Synology>"
endpoint = "http://127.0.0.1:9000"
use_vhost_request = false
use_https = false
\end{verbatim}

\textbf{Nginx federation} --- Matrix federation runs over HTTPS port 443 (not
the default port 8448) using a \texttt{.well-known/matrix/server} file served
by Nginx.  Port 8448 does not need to be open.

\textbf{SSH tunnel restriction} --- the \texttt{tunnel} user on the Oracle VM
is locked to port forwarding only.  The relevant block in
\texttt{/etc/ssh/sshd\_config}:

\begin{verbatim}
Match User tunnel
    AllowTcpForwarding yes
    X11Forwarding no
    PermitTTY no
    PermitOpen 127.0.0.1:9000
\end{verbatim}

\subsection*{Adding a user}

User self-registration is disabled.  To create an account, stop Tuwunel, run
the admin command, then restart:

\begin{verbatim}
systemctl stop tuwunel
tuwunel --config /etc/tuwunel/tuwunel.toml \
    --execute "users create-user <username> <password>"
systemctl start tuwunel
\end{verbatim}

The first user created is automatically granted admin privileges.  Subsequent
users are ordinary users; grant admin via the Element admin panel or the
\texttt{users make-user-admin} command.

\subsection*{Debugging checklist}

\textbf{Tuwunel not responding:}
\begin{verbatim}
systemctl status tuwunel
journalctl -u tuwunel -n 50
\end{verbatim}

\textbf{Media uploads failing / S3 errors:}
\begin{verbatim}
# Check tunnel is up
ss -tlnp | grep 9000
# Check MinIO health
curl http://127.0.0.1:9000/minio/health/live
\end{verbatim}
If port 9000 is not listening, the \texttt{matrix-tunnel} container on the
Synology has dropped.  Log into the Synology via SSH or Container Manager and
restart it:
\begin{verbatim}
docker compose -f /volume1/docker/matrix/docker-compose.yml restart tunnel
\end{verbatim}
If port 9000 is listening but MinIO returns errors, restart the
\texttt{matrix-minio} container the same way.

\textbf{Nginx / TLS errors:}
\begin{verbatim}
nginx -t                  # check config syntax
journalctl -u nginx -n 30
certbot certificates      # check cert expiry
\end{verbatim}

\textbf{TLS certificate renewal} is handled automatically by
\texttt{certbot.timer}.  The certificate expires 2026-09-18; Certbot will
renew it at 60 days before expiry.  If manual renewal is needed:
\begin{verbatim}
systemctl stop nginx
certbot renew --standalone
systemctl start nginx
\end{verbatim}

\textbf{Full health check from outside:}
\begin{verbatim}
curl https://matrix.gull-group.org/.well-known/matrix/server
curl https://matrix.gull-group.org/_matrix/client/versions
\end{verbatim}
The first should return \texttt{\{"m.server":"matrix.gull-group.org:443"\}};
the second should return a JSON list of supported Matrix protocol versions.

\subsection*{Database backup}

Media is stored durably on the Synology via MinIO.  The Tuwunel RocksDB
database at \texttt{/var/lib/tuwunel} on the Oracle VM (user accounts, room
state, message indices) is backed up nightly to the Synology via rsync.

A cron job runs at 03:00 daily on the Synology:

\begin{verbatim}
0  3  *  *  *  root  /volume1/docker/matrix-tunnel/backup.sh \
    >> /var/log/matrix-backup.log 2>&1
\end{verbatim}

The script (\texttt{/volume1/docker/matrix-tunnel/backup.sh}) rsyncs
\texttt{ubuntu@79.76.114.255:/var/lib/tuwunel/} to
\texttt{/volume1/matrix-backup/tuwunel/} using the dedicated key at
\texttt{/volume1/docker/matrix-tunnel/backup\_key}.  For this to work,
\texttt{ubuntu} must be a member of the \texttt{tuwunel} group on the Oracle
VM (\texttt{usermod -aG tuwunel ubuntu}) and \texttt{/var/lib/tuwunel} must
be mode \texttt{0750}.

Check backup logs on the Synology with:
\begin{verbatim}
tail /var/log/matrix-backup.log
\end{verbatim}

\end{document}
